---
title: 'Hands-on: Logistic Regression - Modeling a binary outcome'
subtitle: " LR as a classifier/Predicting Diabetes with a Train/Test Split"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview {.unnumbered}

In this lab we fit a logistic regression and interpret its coefficients.
We will also treat the same kind of model as a **classifier**: we build
it on a **training set**, then judge it on a **test set** it has never
seen. This is the honest way to estimate how a model will perform on new
cases.

**Learning objectives.** By the end of this lab you should be able to:

- Split data into training and test sets reproducibly with `set.seed()`
- Fit a logistic regression on the training set only, interpret the
  results.
- Generate predicted probabilities and class labels for held-out data
- Build a confusion matrix and compute accuracy, sensitivity,
  specificity, and precision
- Run logistic-regression diagnostics: multicollinearity, linearity of
  the logit, influential points, and calibration
- Draw an ROC curve and compute AUC on the test set
- Explain why the decision **threshold** is a choice, not a fixed
  constant

------------------------------------------------------------------------

# Setup & the Data

We'll use the **Pima Indians Diabetes** data from **MASS**. It records
diagnostic measurements for women of Pima heritage, with the goal of
predicting whether each person tested positive for diabetes.

```{r packages}
# install.packages(c("MASS", "tidyverse", "pROC", "car", "broom"))
library(MASS)       # for the Pima diabetes data
library(tidyverse)  # dplyr, ggplot2
library(pROC)       # ROC / AUC
library(car)        # vif() for multicollinearity
library(broom)      # augment() for diagnostics
```

MASS actually ships this data already split into `Pima.tr` (training)
and `Pima.te` (test). For this lab we combine them back into one frame
and then make our **own** split, so you can see exactly how splitting
works. In a real project you would do this once, up front.

```{r load-data}
data(Pima.tr)
data(Pima.te)

pima <- bind_rows(Pima.tr, Pima.te)
glimpse(pima)
```

**Variable key**

| Column  | Meaning                                               |
|---------|-------------------------------------------------------|
| `npreg` | number of pregnancies                                 |
| `glu`   | plasma glucose concentration (oral glucose tolerance) |
| `bp`    | diastolic blood pressure (mm Hg)                      |
| `skin`  | triceps skin-fold thickness (mm)                      |
| `bmi`   | body mass index (kg/m\^2)                             |
| `ped`   | diabetes pedigree function (family-history score)     |
| `age`   | age (years)                                           |
| `type`  | **outcome**: `Yes` = diabetic, `No` = not diabetic    |

```{r outcome-setup}
# Make sure the outcome is a factor with "No" as the reference (first) level,
# so the model predicts the probability of "Yes" (diabetes).
pima <- pima %>% mutate(type = factor(type, levels = c("No", "Yes")))

summary(pima)
```

------------------------------------------------------------------------

# Exploratory Data Analysis

### Outcome balance

Class balance matters: if 90% of patients were "No", a model could score
90% accuracy by ignoring the data and guessing "No" every time. Always
check this before trusting accuracy.

```{r eda-balance}
pima %>%
  count(type) %>%
  mutate(prop = round(n / sum(n), 3))
```

### Predictors vs. the outcome

```{r eda-continuous}
pima %>%
  pivot_longer(c(glu, bmi, age, ped, npreg, bp, skin),
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = type, y = value, fill = type)) +
  geom_boxplot(alpha = 0.6) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Predictor distributions by diabetes status",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r eda-glucose}
ggplot(pima, aes(x = glu, fill = type)) +
  geom_density(alpha = 0.5) +
  labs(title = "Plasma glucose by diabetes status",
       x = "Plasma glucose", y = "Density") +
  theme_minimal()
```

> **Your turn (2 min):** Which predictors separate the two groups most
> clearly? Glucose and BMI are usually strong here - note your guesses
> and compare them to the fitted coefficients later.

------------------------------------------------------------------------

# Train/Test Split

We hold out a random **30%** of the data as a test set and train on the
other **70%**. Two important habits:

- **`set.seed()`** makes the random split reproducible - you (and we)
  get the same split every time the document is knit.
- We split in a **stratified** way (separately within each outcome
  class) so the training and test sets have roughly the same proportion
  of diabetics.

```{r split}
set.seed(2024)

# Stratified 70/30 split: sample row indices within each class.
train_idx <- pima %>%
  mutate(row = row_number()) %>%
  group_by(type) %>%
  slice_sample(prop = 0.70) %>%
  pull(row)

train <- pima[train_idx, ]
test  <- pima[-train_idx, ]

# Check the sizes and that class balance is preserved in both sets.
list(
  n_train = nrow(train),
  n_test  = nrow(test),
  train_balance = round(prop.table(table(train$type)), 3),
  test_balance  = round(prop.table(table(test$type)), 3)
)
```

> **Golden rule:** from here on, the test set is off-limits until the
> very end. All exploring, fitting, and model evaluation happens on
> `train`. Peeking at the test set to make modeling decisions ("data
> leakage") gives you an over-optimistic performance estimate.

------------------------------------------------------------------------

# Train the Model (on training data only)

### Fit the full model

```{r fit-full}
full_model <- glm(
  type ~ npreg + glu + bp + skin + bmi + ped + age,
  data = train,
  family = binomial(link = "logit")
)
summary(full_model)
```

### Simplify with stepwise selection

Model selection is part of "building the model", so it must use the
training data only. We let `step()` drop predictors that don't earn
their place by AIC.

```{r step}
final_model <- step(full_model, direction = "both", trace = 0)
summary(final_model)
```

### Odds ratios for interpretation

```{r odds-ratios}
exp(cbind(OR = coef(final_model), confint(final_model)))
```

Recall the reading: an odds ratio above 1 means higher odds of diabetes
as that predictor increases; below 1 means lower odds. For example, an
OR of about 1.04 for `glu` means each additional unit of plasma glucose
multiplies the odds of diabetes by \~1.04, or increases the odds of
diabetes by 4%, holding the other predictors fixed. If the family
pedigree score increases wih 1 unit, the odds for diabetes will increase
with 2.7 times.

------------------------------------------------------------------------

# Model Diagnostics

A model can have a decent AUC and still violate its assumptions or lean
on a few odd patients. These checks all run on the **training** model -
they are part of building and trusting the model, so they stay on the
training side of the wall.

### Multicollinearity (VIF)

If two predictors carry nearly the same information, their coefficients
become unstable. Variance inflation factors (VIF) flag this. A common
rule of thumb: VIF above 5 is worth a look, above 10 is a real problem.

```{r vif}
vif(final_model)
```

### Linearity of the logit

Logistic regression assumes each **continuous** predictor is linearly
related to the log-odds of the outcome - not to the probability, but to
the logit. We check by plotting each predictor against the model's
linear predictor (the log-odds) with a smoother. Roughly straight lines
support the assumption; strong curves suggest a transformation (e.g. a
log or a quadratic term) might help.

```{r linearity, fig.width=8, fig.height=5}
train %>%
  mutate(logit = predict(final_model, type = "link")) %>%
  pivot_longer(c(npreg, glu, bmi, ped, age),
               names_to = "predictor", values_to = "value") %>%
  ggplot(aes(value, logit)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "#2c7fb8") +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(title = "Linearity of the logit",
       x = NULL, y = "Log-odds (linear predictor)") +
  theme_minimal(base_size = 12)
```

> **Your turn (2 min):** Which predictors look acceptably straight? If
> one curves noticeably, what transformation might you try?

### Influential observations

A single unusual patient shouldn't drive the whole model. **Cook's
distance** measures how much the fitted coefficients would move if an
observation were dropped. A common flag is `4 / n`; points above the
dashed line deserve a closer look (they are candidates for inspection,
not automatic deletion).

```{r influence, fig.width=8, fig.height=4}
diag_df <- augment(final_model) %>% mutate(obs = row_number())
cook_thresh <- 4 / nrow(train)

ggplot(diag_df, aes(obs, .cooksd)) +
  geom_segment(aes(xend = obs, yend = 0), colour = "grey70") +
  geom_hline(yintercept = cook_thresh, linetype = "dashed", colour = "red") +
  labs(title = "Cook's distance", x = "Observation index", y = "Cook's distance") +
  theme_minimal(base_size = 12)

# How many observations exceed the 4/n rule of thumb?
sum(diag_df$.cooksd > cook_thresh)
```

------------------------------------------------------------------------

# Evaluate on the Test Set

Now - and only now - we bring in the held-out test set. We generate
predicted **probabilities** for each test patient, then turn them into
**class labels** using a threshold of 0.5.

```{r predict}
# Predicted probability of "Yes" (diabetes) for each test-set patient.
test_prob <- predict(final_model, newdata = test, type = "response")

# Convert probabilities to class labels. IMPORTANT: give the prediction the
# SAME factor levels as the true outcome, so the confusion-matrix rows and
# columns line up correctly.
test_pred <- factor(
  if_else(test_prob > 0.5, "Yes", "No"),
  levels = levels(test$type)
)
```

### Confusion matrix

```{r confusion}
conf_mat <- table(Predicted = test_pred, Actual = test$type)
conf_mat
```

### Performance metrics

Reading the confusion matrix (with "Yes" = diabetes as the positive
class):

- **TP** true positives, **TN** true negatives
- **FP** false positives (predicted diabetes, actually not)
- **FN** false negatives (missed a real diabetes case)

```{r metrics}
TP <- conf_mat["Yes", "Yes"]
TN <- conf_mat["No",  "No"]
FP <- conf_mat["Yes", "No"]
FN <- conf_mat["No",  "Yes"]

metrics <- tibble(
  accuracy    = (TP + TN) / sum(conf_mat),          # overall correct
  sensitivity = TP / (TP + FN),                     # recall: caught diabetics
  specificity = TN / (TN + FP),                     # correctly cleared non-diabetics
  precision   = TP / (TP + FP)                      # of predicted "Yes", how many were right
)
round(metrics, 3)
```

> **Your turn (3 min):** In a screening setting, which is worse - a
> false positive (telling a healthy person they may have diabetes) or a
> false negative (missing a real case)? How does that shape which metric
> you care about most?

### ROC curve and AUC

The confusion matrix above depends on the 0.5 cutoff. The ROC curve
instead sweeps *every* possible cutoff and plots sensitivity against 1 -
specificity. AUC (area under the curve) summarizes performance across
all thresholds: 0.5 is a coin flip, 1.0 is perfect.

We pull the curve's coordinates out of the `roc` object and draw it in
`ggplot2` so both axes run cleanly from 0 to 1. The dashed diagonal is
the "random guessing" line - a good model bows toward the top-left
corner.

```{r roc, fig.width=6, fig.height=6}
roc_obj <- roc(test$type, test_prob, levels = c("No", "Yes"))

roc_df <- tibble(
  fpr = 1 - roc_obj$specificities,   # false positive rate
  tpr = roc_obj$sensitivities        # true positive rate
) %>% arrange(fpr)

auc_val <- as.numeric(auc(roc_obj))

ggplot(roc_df, aes(fpr, tpr)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey60") +
  geom_path(colour = "#2c7fb8", linewidth = 1.2) +
  annotate("text", x = 0.62, y = 0.10, hjust = 0, size = 4,
           label = paste0("AUC = ", sprintf("%.3f", auc_val))) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  scale_x_continuous(breaks = seq(0, 1, 0.2)) +
  scale_y_continuous(breaks = seq(0, 1, 0.2)) +
  labs(title = "ROC curve (test set)",
       x = "1 - Specificity (False Positive Rate)",
       y = "Sensitivity (True Positive Rate)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
```

## Documentation

### Methods

We analysed the Pima Indians Diabetes data (n = 532 women), in which the
outcome was a binary diagnosis of diabetes (177 positive, 33.3%; 355
negative). The data were partitioned into a training set (70%, n = 371)
and a held-out test set (30%, n = 161) using a stratified random split
with a fixed seed, which preserved the outcome prevalence in both
partitions (training 33.2%, test 33.5%). A logistic regression model
with a logit link was fit on the training set using seven candidate
predictors (number of pregnancies, plasma glucose, diastolic blood
pressure, triceps skin-fold thickness, body mass index, diabetes
pedigree function, and age). The model was simplified by bidirectional
stepwise selection minimising the Akaike Information Criterion (AIC);
all model building was confined to the training set. Model assumptions
were assessed on the training data (multicollinearity via variance
inflation factors, linearity of the logit for continuous predictors, and
influential observations via Cook's distance). Discrimination was
evaluated on the test set using the area under the ROC curve (AUC). A
classification threshold of 0.5 was used for the primary confusion
matrix, with a sensitivity analysis at 0.3.

### Results

Stepwise selection retained five predictors—plasma glucose, body mass
index, diabetes pedigree function, age, and number of
pregnancies—dropping blood pressure and skin-fold thickness; the reduced
model marginally improved on the full model by AIC (328.1 vs 330.2) and
explained a moderate share of outcome variation (McFadden pseudo-R² =
0.33). Higher plasma glucose (odds ratio [OR] 1.04 per unit, 95% CI
1.03–1.05, p \< 0.001), higher BMI (OR 1.09 per unit, 95% CI 1.04–1.13,
p \< 0.001), a higher diabetes pedigree score (OR 2.80, 95% CI
1.17–6.91, p = 0.023), and older age (OR 1.04 per year, 95% CI
1.01–1.08, p = 0.011) were each independently associated with greater
odds of diabetes, whereas number of pregnancies was not statistically
significant (OR 1.09, 95% CI 0.98–1.21, p = 0.10). Multicollinearity was
negligible (all variance inflation factors \< 1.7), and the continuous
predictors were approximately linear on the logit scale. On the held-out
test set the model discriminated well, with an AUC of 0.83 (95% CI
0.77–0.89). At the default 0.5 threshold, overall accuracy was 75.2%,
with specificity of 0.86 and precision of 0.66 but modest sensitivity of
0.54, indicating the model missed nearly half of true diabetes cases at
this cut-off.

------------------------------------------------------------------------

# Thresholds & Wrap-Up

The 0.5 cutoff is a default, not a law. Lowering it flags more patients
as positive - catching more true cases (higher sensitivity) at the cost
of more false alarms (lower specificity). Here's the same test set
scored at a lower threshold of 0.3:

```{r threshold}
test_pred_30 <- factor(
  if_else(test_prob > 0.6, "Yes", "No"),
  levels = levels(test$type)
)
cm30 <- table(Predicted = test_pred_30, Actual = test$type)
cm30

tibble(
  threshold   = 0.60,
  sensitivity = cm30["Yes", "Yes"] / sum(cm30[, "Yes"]),
  specificity = cm30["No",  "No"]  / sum(cm30[, "No"])
) %>% round(3)
```

Compare these numbers to the 0.5 results above and notice the trade-off.

## Wrap-up exercises

Try these on your own:

1.  Re-run the whole notebook with a different seed (say `set.seed(1)`).
    Do the test-set accuracy and AUC change a lot? What does that tell
    you about trusting a single split?
2.  Sweep thresholds from 0.1 to 0.9 and plot sensitivity and
    specificity against the threshold. Where do the two curves cross?
3.  Fit a model using **only** `glu` and `bmi`. How much test-set AUC do
    you lose compared to the full model? Is the simpler model "good
    enough"?
4.  Split the data 50/50 instead of 70/30. Does the smaller training set
    hurt test performance? Why might it?
5.  In one short paragraph, report your final model's test-set
    performance the way you would to a clinician: accuracy, how many
    real cases it catches, and how often it raises a false alarm.

```{r session-info}
sessionInfo()
```
